---
title: "Kennewick PCA"
author: "Christian Porras"
date: "2023-10-17"
output: html_document
---

# Ancestry inference PCA

## Visualizing genetic relatedness

```{r}
package_names <- c("tidyverse","plotly","ggplot2")

# load packages, install them first if not installed
for (package_name in package_names){
  if (!requireNamespace(package_name, quietly = TRUE)) {
    install.packages(package_name)
  }
  
  library(package_name, character.only = TRUE)

  }
```

```{r}
### Load in data files
# check working directory 
setwd("~/Desktop/creigs_ta_2023/kennewick-pca")

df = read_csv('data/1K_genomes_PCA.csv')
PC_load = read_csv('data/1kGenomesPCloading.csv')
superpop_key = read_tsv("data/20131219.superpopulations.tsv")
populations = read_tsv('data/20131219.populations.tsv')
pop_key = populations[,1:3]
```

### Glimpse at the data set we'll be studying
```{r}
glimpse(df)
```
> We want to infer the population that our first sample was chosen from by comparing genetic data across a global sample from the [1000 Genomes Project](https://en.wikipedia.org/wiki/1000_Genomes_Project).


### Unlabeled PCA plot
```{r}
pc1 = round(PC_load$loading[1]*100)
pc2 = round(PC_load$loading[2]*100)

plot(df$PC1_0.03008,
     df$PC2_0.01143,
     main = '1K Genomes test PCA plot',
     xlab = paste("PC1 ",pc1,"% variance explained"),
     ylab = paste("PC2 ",pc2,"% variance explained")
     )

```

> We can plot the first two principal components for each sample, but this doesn't tell us much without labels!

```{r}
################# Legend key for Population codes #################
superpop_key
```

### Label PCA plot by super population

```{r}
library(ggplot2)

pc1 = round(PC_load$loading[1]*100)
pc2 = round(PC_load$loading[2]*100)

# Set up the plot
ggplot(data = df, aes(x = PC1_0.03008, y = PC2_0.01143)) +
  geom_point(aes(color = `Super_Population`), size = 4, alpha = 0.2) +
  geom_point(data = subset(df, Super_Population == 'UNKNOWN'), color = 'black', size = 4, shape = 15) +
  scale_color_discrete(name = 'Super Population') +
  labs(
    title = '1K Genomes PCA by super population',
    x = paste("PC1 ",pc1,"% variance explained"),
    y = paste("PC2 ",pc2,"% variance explained")
  ) +
  theme_minimal() +
  theme(legend.position = 'right') +
  theme(legend.text = element_text(size = 18)) +
  guides(color = guide_legend(override.aes = list(size = 5))
  )

```

> Now we're getting somewhere! Still, we can use another dimension to better view our unknown.

```{r}
pc1 = round(PC_load$loading[1]*100)
pc2 = round(PC_load$loading[2]*100)
pc3 = round(PC_load$loading[3]*100)

# Create the ggplot plots for each PC combination
plot1 <- ggplot(data = df, aes(x = PC1_0.03008, y = PC2_0.01143)) +
  geom_point(aes(color = `Super_Population`), size = 4, alpha = 0.2) +
  geom_point(data = subset(df, Super_Population == 'UNKNOWN'), color = 'black', size = 4, shape = 15) +
  scale_color_discrete(name = 'Super Population') +
  labs(
    title = 'PC1 vs PC2',
    x = paste("PC1 ",pc1,"% variance explained"),
    y = paste("PC2 ",pc2,"% variance explained")
  ) +
  theme_minimal() +
  theme(legend.position = 'right') +
  theme(legend.text = element_text(size = 18)) +
  guides(color = guide_legend(override.aes = list(size = 5))
  )

plot2 <- ggplot(data = df, aes(x = PC1_0.03008, y = PC3_0.00354)) +
  geom_point(aes(color = `Super_Population`), size = 4, alpha = 0.2) +
  geom_point(data = subset(df, Super_Population == 'UNKNOWN'), color = 'black', size = 4, shape = 15) +
  scale_color_discrete(name = 'Super Population') +
  labs(
    title = 'PC1 vs PC3',
    x = paste("PC1 ",pc1,"% variance explained"),
    y = paste("PC3 ",pc3,"% variance explained")
  ) +
  theme_minimal() +
  theme(legend.position = 'right') +
  theme(legend.text = element_text(size = 18)) +
  guides(color = guide_legend(override.aes = list(size = 5))
  )

plot3 <- ggplot(data = df, aes(x = PC2_0.01143, y = PC3_0.00354)) +
  geom_point(aes(color = `Super_Population`), size = 4, alpha = 0.2) +
  geom_point(data = subset(df, Super_Population == 'UNKNOWN'), color = 'black', size = 4, shape = 15) +
  scale_color_discrete(name = 'Super Population') +
  labs(
    title = 'PC2 vs PC3',
    x = paste("PC2 ",pc2,"% variance explained"),
    y = paste("PC3 ",pc3,"% variance explained")
  ) +
  theme_minimal() +
  theme(legend.position = 'right') +
  theme(legend.text = element_text(size = 18)) +
  guides(color = guide_legend(override.aes = list(size = 5))
  )

plot1
plot2
plot3
```
> If only we could plot in 3D and rotate our plot around to see different angles...

```{r}
################ Legend key describing sub population codes ################
pop_key
```


```{r}
library(plotly)

plot_ly(df, x = ~PC1_0.03008, 
        y = ~PC2_0.01143, 
        z = ~PC3_0.00354, 
        color = ~Super_Population, type = "scatter3d")

```

> Try hiding various combinations of subpopulations to identify which of these cluster most closely to our unknown. 

```{r}
library(plotly)

plot_ly(df, x = ~PC1_0.03008, 
        y = ~PC2_0.01143, 
        z = ~PC3_0.00354, 
        color = ~Super_Population, 
        symbol = ~Population,
        type = "scatter3d")

```

Which ancestry group is our unknown sample most similar to in PCA?
